```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```


## Download PFTC data

We have created a R package to download the data, called *dataDownloader*.
If you are using the package for the first time you need to install the package using the  command below.
If you have used it before, just run the second line of code to load the package.

```{r, package, eval=FALSE}
devtools::install_github("Between-the-Fjords/dataDownloader")
library(dataDownloader)
```

Now you can download the all the files you need.
Let's download the community data and the trait from the Svalbard course.

```{r, download-svalbard, eval=FALSE}
get_file(node = "smbqh",
         file = "PFTC4_Svalbard_2018_Gradient_Traits.csv",
         path = "webpage/data",
         remote_path = "Traits")
                       
get_file(node = "smbqh",
         file = "PFTC4_Svalbard_2018_Community_Gradient.csv",
         path = "webpage/data",
         remote_path = "Community")
                       
```


### Exercise {- .exercise .toc-ignore}

Now it is your turn.
Copy the code into your console and download the data.

